Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-27 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision on the final part of section 2 and the beginning of section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2901 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ecc1143bfc3733cc8672ddd486aa0b8fc102f3bb

In already existing generalized linear or additive approaches to circular regression modeling 
usually the location parameter $\mu$ is regressed on covariates in the general form 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} z).
\end{equation}
with a circular intercept $\mu_0$, a vector of covariates $z$
with the corresponding vector of coefficients $\beta$ that are to be estimated,
and a link function $g$.
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an interval of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
constant value. Additionally to these two model variations where only one of the two distribution 
parameters depends on covariates a combination of them is introduced in a joint model. There, the 
two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
in alternation within an iterative optimization algorithm. 
Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
more general case by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
However, in practice the incorporated optimization procedure has shown to be very complex, already
for only one parameter being regressed on covariates and consequently even more so in the combined model.
This is mainly due to characteristics of the log-likelihood function of the employed von Mises distribution
with the plugged-in link function. In particular, the function is not globally concave and possibly very 
sharp local maxima can prevent the algorithm from converging.
Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
true underlying effect is an abrupt shift.
%% FIXME: difficulties in modeling smooth effects on the unit circle
One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
is to employ tree-based models as presented in the following Section.

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}
%% FIXME: what are we aiming at? which problems do we want to avoid/solve?
%% mu and kappa as constants in subgroups, no initial values
%% Lund: what do they present (including formulas)
%% compare, what is different, how do we improve this approach

A common way of including covariates is to regress the model parameters on
covariates. In the case of a probabilistic model for a circular response this
corresponds to expressing the location parameter $\mu$ and the concentration
parameter $\kappa$ of the von Mises distribution by means of the available
covariates, e.g., in the form of a linear combination, possibly with an
appropriate link function plugged in as in GLMs. This kind of approach is
presented for example in the mixed model of \cite{Fisher+Lee:1992} and
\cite{Fisher:1993}. Apart from the methodology developed specifically for
circular data more general approaches to regressing model parameters on
covariates are commonly used such as for example generalized additive models
for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
where each distribution parameter is estimated by its own generalized additive
model (GAM). However, difficulties might arise in case of non-additive changes
and a proper model specification can be very challenging, particularly for a
high number of covariates and no information on possible interactions. As an
alternative regression approach which can deal with these problems, we
introduce a circular probabilistic tree-structured model in this section.

>>>>

In already existing generalized linear or additive approaches to circular regression modeling 
usually the location parameter $\mu$ is regressed on covariates in the general form 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} z).
\end{equation}
with a circular intercept $\mu_0$, a vector of covariates $z$
with the corresponding vector of coefficients $\beta$ that are to be estimated,
and a link function $g$.
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an interval of length $2\pi$ by setting 
$g(x) = 2\pi \cdot F(x)$ 
for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
constant value. Additionally to these two model variations where only one of the two distribution 
parameters depends on covariates a combination of them is introduced. In this joint model the 
two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
in alternation within an iterative optimization algorithm. 
Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
more general case by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
However, in practice the incorporated optimization procedure has shown to be very complex, already
for only one parameter being regressed on covariates and consequently even more so in the joint model.
This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
distribution with the plugged-in link function. In particular, the function is not globally concave and
possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
compression of the estimated density function.
\newline
Referring to additive models in general and hence also to their application on circular data, it has to
be considered that a proper model specification can be very challenging, particularly for a high number 
of covariates and no information on possible interactions. Moreover, the additive structure might impose 
a smooth effect even  if the true underlying effect is an abrupt shift.
\newline
One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
is to employ tree-based models as presented in the following Section.

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}
The main goal of employing tree-based models to circular data is to provide a flexible and 
easy-to-apply modeling approach which allows to detect additive as well as abrupt changes while
accounting for a proper representation on the unit circle. 
%A very flexible framework should allow for detecting a wide range of 
%effects, from abrupt changes to smooth transitions, while for the latter difficulties related
%to a circular representation should be avoided as well as any dependence on initial values.
\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
and where to split the data an angular distance measure is evaluated between the mean direction 
and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
Hence, only the location parameter is considered for splitting such that effects in other parameters 
might be missed. Therefore, we introduce a distributional approach to circular regression trees 
where a fully specified distributional model is fitted in each node of the tree which enables the 
detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
quantities.
Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
within each node is fitted without employing covariates.
%A common way of including covariates is to regress the model parameters on
%covariates. In the case of a probabilistic model for a circular response this
%corresponds to expressing the location parameter $\mu$ and the concentration
%parameter $\kappa$ of the von Mises distribution by means of the available
%covariates, e.g., in the form of a linear combination, possibly with an
%appropriate link function plugged in as in GLMs. This kind of approach is
%presented for example in the mixed model of \cite{Fisher+Lee:1992} and
%\cite{Fisher:1993}. Apart from the methodology developed specifically for
%circular data more general approaches to regressing model parameters on
%covariates are commonly used such as for example generalized additive models
%for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
%where each distribution parameter is estimated by its own generalized additive
%model (GAM). However, difficulties might arise in case of non-additive changes
%and a proper model specification can be very challenging, particularly for a
%high number of covariates and no information on possible interactions. As an
%alternative regression approach which can deal with these problems, we
%introduce a circular probabilistic tree-structured model in this section.
